Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-06-07 Manuscript/Manuscript.tex
Ms: Methods edit, key resources table update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
  onecolumn]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Arial}
  \setmonofont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance in zooplankton},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates
UV avoidance in zooplankton}
\author{true}
\date{}

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key
regulator of animal physiology. Here we uncover a role for NO in shaping
circuit dynamics to orchestrate light-avoidance behaviour. We studied
UV-avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval
\emph{Platynereis dumerilii}, a marine annelid. We found NOS
specifically expressed in interneurons (INNOS) postsynaptic to cPRCs.
Stimulation of cPRCs by violet light leads to cPRC inhibition but
concomitant INNOS activation and NO production. NO feeds back to cPRCs
and triggers their delayed activation through an unconventional guanylyl
cyclase. This results in the activation of projection interneurons and
the inhibition of serotonergic ciliomotor neurons. In \emph{NOS}
mutants, NO feedback and projection-neuron activation do not occur and
avoidance behaviour is defective. By mathematical modelling, we could
capture the circuit dynamics in both wild-type and mutant larvae. Our
results reveal how NO-mediated neuroendocrine signalling gates a
synaptic circuit to initiate light-avoidance behaviour.

@@

Neuronally produced nitric oxide (NO) can act as a diffusible
neuromodulator. This small free-radical molecule is synthesized by
nitric oxide synthase (NOS)\ldots{}

In the mammalian retina, NO is produced by amacrine cells and activates
a \emph{soluble guanylate cyclase (sGC)} in cone bipolar cells. The
NO-dependent increased in cGMP tunes bipolar cell activity \ldots.
{[}refs{]} {[}other examples{]}

To investigate NO signalling in the modulation of neural circuit
dynamics, we studied the larval stages of the marine annelid model
\emph{Platynereis dumerilii} (Ozpolat et al., 2021). We uncovered an
essential function for NO in these larvae in UV/violet-light avoidance.
In \emph{Platynereis} larvae, UV/violet avoidance is characterised by
downward swimming and is mediated by the UV/violet-responding brain
ciliary photoreceptor cells (cPRCs).

The cPRCs express a vertebrate-type ciliary opsin, c-opsin1 (Arendt et
al., 2004). This opsin forms a UV-absorbing bistable photopigment with
an absorption maximum around 384 nm Verasztó et al. (2018).
\emph{c-opsin1} knockout larvae show defective UV/violet avoidance
(Verasztó et al., 2018). Here we show that \emph{Platynereis} \emph{NOS}
is expressed in interneurons of the cPRC circuitry and is required for
UV/violet-avoidance behaviour. By combining calcium imaging across the
fully-mapped cPRC circuit, genetic perturbations and mathematical
modelling, we describe how NO tunes circuit dynamics through
non-synaptic reverse signalling to cPRCs. This delayed neuroendocrine
feedback is essential for circuit output, and integrates UV/violet input
and also induces a short-term memory manifest in altered circuit
activity.

\#In adult worms, UV light and c-opsin1 modulate locomotion and
maturation (Veedin Rajan et al., 2021).

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of metazoan NOS sequences
\textbf{(Figure 1---figure supplement 1)}. To characterise the
expression pattern of \emph{NOS} we used in situ hybridization chain
reaction (HCR) and transient transgenesis. In two- and three-day-old
larvae, we detected \emph{NOS} expression in four cells (two of them
weakly expressing) in the apical organ region \textbf{(Figure 1D and
Figure 1---figure supplement 2)}. \emph{NOS} was also expressed in the
region of the visual eyes (adult eyes) and the pigmented eyespots
\textbf{(Figure 1---figure supplement 1)}. The four apical organ cells,
but not the eyes, were also labelled with a \emph{NOS} reporter
construct driving palmitoylated-tdTomato \textbf{(Figure 1E)}. This
reporter also revealed the axonal projections of these central
NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017)
\textbf{(Figure 1B,C)}. In the synaptic connectome, the INNOS cells are
postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW
interneurons, which are also cPRC targets \textbf{(Figure 1C,F)}. INRGW
neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1),
which synapse on the prototroch ciliary band and the cholinergic MC
ciliomotor neurons \textbf{(Figure 1C,F)}.

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** SEM image of a three-day-old Platynereis larva. **(B, C)** Five types of neurons (cPRC, INNOS, INRGW, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a three-day-old larva. Ciliated cells labelled with (B) are shown in grey. The INNOS on the ventral side labelled with (C) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. **(D)** Expression analysis of the NOS gene (magenta) using in situ HCR. Anterior view of the larva at two-day-old. Co-stained image using acetylated α-tubulin antibody (acTub: green). **(E)** Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (NOSp::palmi-3xHA: magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia of cPRC and axonal scaffold (acTub: green). Views of larvae at three-day-old from the anterior side. Insert: One of the reconstructed INNOS is shown from the anterior side. **(F)** Wiring diagram of the cPRC circuit. Hexagons and arrows represent cell groups and their synaptic connections, respectively. Numbers inside hexagons indicate the number of cells grouped in each node. Arrow line thickness (synapse weight) is equal to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

\textbf{Nitric oxide produced during UV/violet stimulation gates the
output of the cPRC circuit}

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the ramified sensory cilia of the cPRCs to 405 nm violet light.
Following light stimulation of the cPRCs but not a control area, we
detected an increase in DAF-FM fluorescence in the anterior
neurosecretory neuropil, the region of INNOS projections \textbf{(Figure
2)}.

\begin{figure}
\includegraphics[width=29.17in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** A NO-positive cell visualised by DAF-FM (green) is shown (dashline). The white line indicates the outer frame of the larva. All subsequent images are of two-day-old larvae. **(B)** Intensity changes before (upper) and during 405 nm light stimulation (lower) in a cell visualised by DAF-FM. **(C)** The intensity of DAF-FM changes over time when cPRC (green) and other cells (ctr: gray) are stimulated with UV/violet (405 nm) light. Purple squares indicate the timing of 405 nm light stimulation, the transparent lines show the optical responses to individual intensities, each intensities are normalized (ΔF/F0) and the dark trace lines and the shaded bands show the mean value and the standard error range of the mean.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

Next, we wanted to directly test whether NO signalling is involved in
UV/violet avoidance. To achieve this, we generated two
\emph{Platynereis} \emph{NOS} knockout lines with the CRISPR/Cas9
method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both
leading to frame-shift and an early stop codon and thus likely
representing null alleles \textbf{(Figure 3A)}. We could establish a
homozygous line for both mutations indicating that NOS is not an
essential gene. To quantify UV avoidance, we recorded the trajectories
of freely swimming wild type and mutant larvae in vertical columns,
illuminated laterally from two opposite sides with 395 nm UV light
\textbf{(Figure 3B, C and Figure 3---figure supplement 1A-D)}. As
previously shown, wild-type larvae swim downward following
non-directional UV/violet light stimulation (Verasztó et al., 2018). In
contrast, both two- and three-day-old homozygous \emph{NOS}-mutant
larvae showed a strongly diminished UV-avoidance response
\textbf{(Figure 3B-D and Figure 3---figure supplement 1)}. This
phenotype is similar to the defective UV-avoidance of \emph{c-opsin1}
mutant larvae and reveals a requirement for NOS in UV-avoidance
behaviour (Verasztó et al., 2018). Wild type but not mutant larvae also
showed an increase in swimming speed under UV light that may be due to
downward swimming trajectories (swimming with gravity) \textbf{(Figure
3---figure supplement 1C,D)}. We also tested directional phototaxis, by
exposing larvae to 480 nm directional and collimated light from the top
of the column. Three-day-old but not two-day-old \emph{NOS}-mutant
larvae also showed reduced phototactic behaviour, suggesting a function
for \emph{NOS} in the visual eyes that mediate three-day-old phototaxis
(Randel et al., 2014) \textbf{(Figure 1---figure supplement 1A,B)}. To
distinguish between an acute or developmental function of NOS in light
responses, we next tested larvae exposed to the NOS inhibitor L-NAME.
Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance. In contrast, phototaxis was
not effected \textbf{(Figure 3E)}. Overall, our results indicate an
acute requirement for NOS signalling in UV-avoidance and a possible
indirect, developmental role in visual phototaxis.

\begin{figure}
\includegraphics[width=33.33in]{figures/Fig3} \caption{**Figure 3. Larvae with inhibited NOS have a delayed response to UV-avoidance.** **(A)** Top: The domain and Exon/Intron structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Gray shows PAM sequences, red shows stop codons. **(B)** Overlaid trajectories for WT (n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) at three-day-old larvae. 0 sec as the starting point. After 10 sec, UV (395 nm) stimulation from the side. **(C)** The temporal changes in the vertical position of the WT and mutant larvae before (dark) and after UV stimulation (side 395 nm) are shown. The starting points of each larval trajectory are set to 0. **(D)** Vertical swimming in wild-type (WT) and mutant (NOSΔ11 and NOSΔ23) larvae at three-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins. **(E)** Vertical swimming in larvae treated with NOS inhibitors, L-NAME at three-day-old stimulated with UV (395 nm) light from side, blue (488 nm) light from top and UV (395 nm) light from top. The data are shown in 30 s bins.}\label{fig:unnamed-chunk-3}
\end{figure}

\textbf{NO feedback tunes cPRC responses to UV/violet stimulation}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  10pt,
  onecolumn]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Arial}
  \setmonofont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.5in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Nitric oxide feedback to ciliary photoreceptor cells gates a UV
avoidance circuit}
\author{true}
\date{}

@@

Nitric oxide (NO) produced by nitric-oxide synthase (NOS) is a key
regulator of animal physiology. Here we uncover a function for NO in the
integration of UV exposure and the gating of a UV-avoidance circuit. We
studied UV/violet avoidance mediated by brain ciliary photoreceptors
(cPRCs) in larvae of the annelid \emph{Platynereis dumerilii}. In the
larva, NOS is expressed in interneurons (INNOS) postsynaptic to cPRCs.
UV stimulation of cPRCs triggers INNOS activation and NO production. NO
signals retrogradely to cPRCs to induce their sustained post-stimulus
activation through an unconventional guanylyl cyclase. This late
activation inhibits serotonergic ciliomotor neurons to induce downward
swimming. In \emph{NOS} mutants, retrograde signalling, circuit output
and UV avoidance are defective. By mathematical modelling, we
recapitulate phototransduction and circuit dynamics in wild-type and
mutant larvae. Our results reveal how NO-mediated retrograde signalling
gates a synaptic circuit and induces short-term memory of UV exposure to
orchestrate light-avoidance behaviour.

@@

In nervous systems, synaptic transmission and volume transmission
together shape circuit dynamics (Bargmann and Marder, 2013). While
synaptic transmission occurs at specialised contact sites, volume
transmission is characterised by the delocalised release of diverse
diffusive neuromodulators.

Nitric oxide (NO) is one such modulator with unique physical and
signalling properties. This free radical synthesized from L-arginine by
nitric oxide synthase (NOS) is short-lived and can diffuse across
biological membranes (Cudeiro and Rivadulla, 1999; Thomas, 2015).
Canonical NO signalling involves the Ca2+/calmodulin-dependent
activation of NOS, NO production and diffusion, and the NO-dependent
activation of soluble guanylate cyclases (sGC). This can occur
cell-autonomously or in other cells leading to cGMP production (Bredt et
al., 1990; Hölscher, 1997). Given that NOS activation is calcium
dependent and NOS shows neuron-type-specific expression (Aso et al.,
2019; Gibbs and Truman, 1998; Mobley et al., 2022; Wildemann and Bicker,
1999), NO action can lead to the activity-dependent modulation of neural
circuits at specific sites (Aso et al., 2019; Jacoby et al., 2018;
Vielma et al., 2014; Wang et al., 2007) .

In the vertebrate retina, NOS is expressed in amacrine, ganglion and
other cells and the actions of NO can be diverse (Cudeiro and Rivadulla,
1999; Jacoby et al., 2018; Wang et al., 2007). For example, defective NO
signalling in NOS knockouts leads to a decreased sensitivity of retinal
ganglion cells to light stimulation (Wang et al., 2007). In the retina,
NO signalling can also involve pathways other than canonical sGC
signalling (Jacoby et al., 2018; Tooker et al., 2013; Wei et al., 2012).
Due to the complex expression of NOS in vertebrates and the diversity of
its functions, it has been challenging to link the neurophysiological
effects of NO to signalling mechanisms and behaviour change.

In the \emph{Drosophila} brain, NO signalling is also involved in tuning
circuit activity at diverse and specific sites. In the ellipsoid body of
the central complex, NO is involved in visual working memory (Kuntz et
al., 2017). NO signalling also tunes the dynamics of associative
memories in mushroom body circuits. NOS is expressed in PPL1-γ1pedc
mushroom-body neurons where it is involved in shortening memory
retention while promoting fast memory updating in response to new
experiences (Aso et al., 2019).

In the whole organism context, NO signalling has often been studied in
diverse marine invertebrates, where NO can regulate larval settlement
and metamorphosis (Leise et al., 2001; Locascio et al., 2022; Song et
al., 2021; Ueda et al., 2016; Zhang et al., 2012). While NOS-expressing
neurons potentially responsible for these effects have been reported
(Bishop and Brandhorst, 2007; Locascio et al., 2022), it has not been
possible to link these NO-dependent effects on behaviour or life-cycle
transitions to neuronal activity and function.

Overall, we still know little about how NO production relates to
stimulus conditions, how it shapes circuit activity at specific neuron
types, and how NO-dependent modulation relates to behaviour.

To investigate NO function in neural circuit dynamics and behaviour,
here we study larvae of the marine annelid model \emph{Platynereis
dumerilii} (Ozpolat et al., 2021). \emph{Platynereis} has emerged as a
model for systems neuroscience with a toolkit that enables the
combination of behavioural analysis and neuronal activity imaging with
genetic manipulations. A whole-body synaptic connectome and gene
expression atlases are also available and can be integrated with
functional approaches in live animals thanks to the cellular-level
stereotypy of larvae of the same developmental stage (Ozpolat et al.,
2021; Verasztó et al., 2020; Verasztó et al., 2017; Vergara et al.,
2021).

Here we uncovered an essential function for NO signalling in larval
UV/violet-light avoidance behaviour. In \emph{Platynereis} larvae,
UV/violet avoidance is mediated by brain ciliary photoreceptor cells
(cPRCs) and is characterised by downward swimming (Verasztó et al.,
2018). The cPRCs express a ciliary-type opsin, c-opsin1 (Arendt et al.,
2004) that forms a UV-absorbing bistable photopigment with an absorption
maximum around 384 nm (Tsukamoto et al., 2017; Veedin Rajan et al.,
2021; Verasztó et al., 2018). Upon UV/violet exposure, the cPRC show a
characteristic biphasic calcium response that is \emph{c-opsin1}
dependent. UV/violet avoidance is also \emph{c-opsin1}-dependent and is
defective in \emph{c-opsin1} mutants (Verasztó et al., 2018). Here we
show that \emph{Platynereis} \emph{NOS} is expressed in interneurons of
the cPRC circuit and is required for UV/violet-avoidance. By combining
calcium imaging across the fully-mapped cPRC circuit (Verasztó et al.,
2018) with genetic perturbations and mathematical modelling, we describe
how NO tunes circuit dynamics through non-synaptic retrograde signalling
to cPRCs. This delayed neuroendocrine feedback integrates UV/violet
exposure to induce a short-term memory manifested in altered circuit
activity and an aversive behavioural response.

@@

We identified a single \emph{nitric oxide synthase} (\emph{NOS}) gene in
the \emph{Platynereis dumerilii} genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that \emph{Platynereis}
NOS belongs to an orthology group of bilaterian NOS sequences (Figure
1---figure supplement 1). To characterise the expression pattern of
\emph{NOS} we used in situ hybridization chain reaction (HCR) and
transient transgenesis. In two- and three-day-old larvae, we detected
\emph{NOS} expression in four cells (two of them weakly expressing) in
the apical organ region (Figure 1D and Figure 1---figure supplement 2).
\emph{NOS} was also expressed in the region of the visual eyes (adult
eyes) and the pigmented eyespots (Figure 1---figure supplement 2). The
four apical organ cells, but not the eyes, were also labelled with a
\emph{NOS} reporter construct driving palmitoylated-tdTomato (Figure
1E). This reporter also revealed the axonal projections of these central
NOS-expressing neurons. The position and morphology of the four
\emph{NOS+} cells allowed us to identify the same four cells as four
interneurons (INNOS) in our three-day-old whole-body \emph{Platynereis}
volume EM data (Verasztó et al., 2020; Williams et al., 2017) (Figure
1B,C). In the synaptic connectome, the INNOS cells are postsynaptic to
the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which
are also cPRC targets (Figure 1C,F). The projections of INNOS cells are
segregated into input (dendritic) and output (axonal) compartments, with
cPRC inputs occurring on the dendritic part in the dense neurosecretory
plexus. INNOS output synapses form in the more ventral projection region
(Figure 1---figure supplement 3). INRGWa neurons synapse on the head
serotonergic ciliomotor neurons (Ser-h1), which synapse on the
prototroch ciliary band and the cholinergic MC ciliomotor neurons
(Figure 1C,F) (Verasztó et al., 2017).

\begin{figure}
\includegraphics[width=25in]{figures/Fig1} \caption{**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** **(A)** Scanning electron microscopy image of a three-day-old *Platynereis* larva. **(B, C)** Volume rendering of the neuron types (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed from a whole-body transmission electron microscopy volume of a three-day-old larva. Neurite skeletons are shown with cell-body positions represented by spheres. Projections of all neurons in the body are shown in grey to highlight the neuropils. The outline of the yolk is also indicated in grey. In B, nuclei positions of the prototroch head ciliary band are shown as grey spheres. **(D)** Expression of the *NOS* gene detected by in situ HCR (magenta) in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(E)** Expression of a *NOS* reporter (NOSp::palmi-3xHA: magenta) labelled with an anti-HA antibody in a two-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil. **(F)** Synaptic wiring diagram of the cPRC circuit. Hexagons represent cell groups, with the number of cells per group shown in square brackets. Arrows represent the summed number of synaptic contacts between cell groups. Arrow thickness is proportional to the number of synapses.}\label{fig:unnamed-chunk-1}
\end{figure}

\textbf{Nitric oxide is produced during UV/violet stimulation of the
cPRCs}

The expression of \emph{NOS} in the INNOS interneurons in the cPRC
circuit suggests that NO signalling may be involved in
UV/violet-avoidance. To test this, first we asked whether NO is produced
during UV/violet stimulation of the larvae. We injected the fluorescent
NO-reporter DAF-FM into zygotes and imaged two-day-old larvae while
exposing the region of cPRC cilia to 405 nm violet light. To mark cell
outlines, we coinjected mRNA encoding a red-fluorescent reporter
(RGECO), allowing the identification of the cPRCs. Following light
stimulation in the region of the ramified cilia of the cPRCs, we
detected an increase in DAF-FM fluorescence in the anterior
neurosecretory neuropil, the region of INNOS projections. This increase
did not occur in larvae where we illuminated a control area (Figure 2).

\begin{figure}
\includegraphics[width=29.17in]{figures/Fig2} \caption{**Figure 2. NO produced by UV/violet stimulation to cPRCs.** **(A)** DAF-FM fluorescence in the region of the neurosecretory neuropil. The white line indicates the outline of the larva. The dashed line corresponds to the area where fluorescence was quantified. Circles indicate the location of cPRC and control stimulation. cPRCs are marked by thin lines. **(B)** Changes in DAF-FM fluorescence before and during 405 nm light stimulation. **(C)** Changes in DAF-FM fluorescence over time during 405 nm stimulation of the cPRCs (green) or a control area (ctr: gray). The purple box indicates the duration of 405 nm stimulation. Individual traces normalized (ΔF/F0) are shown as thin lines. Thick lines show the mean value with 0.95 confidence intervals. N = 9 larvae for control and 11 for cPRC stimulation.}\label{fig:unnamed-chunk-2}
\end{figure}

@@

We next tested whether NO signalling regulates UV/violet avoidance. To
achieve this, we generated two \emph{Platynereis} \emph{NOS} knockout
lines with the CRISPR/Cas9 method. We recovered two deletions
(NOSΔ11/Δ11 and Δ23/Δ23), both frame-shift mutations leading to an early
stop codon and thus likely representing null alleles (Figure 3---figure
supplement 1A). We could establish a homozygous line for both mutations
indicating that NOS is not an essential gene in \emph{Platynereis}. To
quantify UV avoidance, we recorded the trajectories of freely swimming
wild type and mutant larvae in vertical columns, illuminated laterally
from two opposite sides with 395 nm UV light (Figure 3A and Figure
3---figure supplement 1B). As previously shown, wild-type larvae swim
downward following non-directional UV/violet light stimulation (Verasztó
et al., 2018). In contrast, both two- and three-day-old homozygous
\emph{NOS}-mutant larvae showed a strongly diminished UV-avoidance
response (Figure 3A, B and Figure 3---figure supplement 1B,C). This
phenotype is similar to the defective UV-avoidance of \emph{c-opsin1}
mutant larvae (Verasztó et al., 2018) and reveals a requirement for NOS
in UV-avoidance behaviour. Wild type but not mutant larvae also showed
an increase in swimming speed under UV light that may be due to downward
swimming trajectories (swimming in the direction of gravity) (Figure 3B
and Figure 3---figure supplement 1C). We also tested directional
phototaxis, by exposing larvae to 480 nm directional collimated light
from the top of the column. Three-day-old but not two-day-old
\emph{NOS}-mutant larvae also showed reduced phototactic behaviour,
suggesting a function for \emph{NOS} in the visual eyes that mediate
three-day-old phototaxis (Randel et al., 2014) (Figure 3D and Figure
3---figure supplement 1G).

To distinguish between an acute and developmental function of NOS in
light responses, we next tested larvae exposed to the NOS inhibitor
L-NAME. Larvae incubated for 5 min in 0.1 mM or 1 mM L-NAME showed a
dose-dependent inhibition of UV avoidance. In contrast, phototaxis was
not affected (Figure 3C, E). Overall, our results indicate an acute
requirement for NOS signalling in UV-avoidance and a possible indirect,
developmental role in the visual system, reminiscent of the function of
NO signalling in \emph{Drosophila} eye development (Gibbs and Truman,
1998).

\begin{figure}
\includegraphics[width=33.33in]{figures/Fig3} \caption{**Figure 3. NOS is required for UV avoidance in *Platynereis* larvae.**  **(B)** Swimming trajectories of wild type (WT, n=32) and NOS mutant (NOSΔ11/Δ11, n=26 and NOSΔ23/Δ23, n=47) three-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec after the onset of 395 nm stimulation from the side. **(B)** Vertical position of batches of wild type and mutant three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(C)** Vertical position of batches of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. **(D)** Vertical displacement in 30 sec bins of wild type and mutant (NOSΔ11 and NOSΔ23) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. **(E)** Vertical displacement in 30 sec bins of control and L-NAME-treated (0.1 and 1 mM) three-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top.}\label{fig:unnamed-chunk-3}
\end{figure}

\textbf{NO retrograde signalling tunes cPRC responses to UV/violet
stimulation}

To investigate how NO signalling alters the dynamics of the cPRC
circuit, we carried out calcium imaging experiments. We ubiquitously
expressed the calcium sensor GCaMP6s in larvae and imaged calcium
signals during 405 nm stimulation of the cPRCs. As we have shown
previously, a 20-sec local stimulation of cPRC cilia led to a transient
increase in cPRC calcium levels, followed by a transient decrease
(Verasztó et al., 2018). After \textasciitilde20-sec, calcium levels in
cPRCs were raising again, reaching higher levels than at the start of
the stimulus -- a response that may involve depolarisation (Figure 4A,
B). This activation phase occurs after the 20 sec stimulation period and
is likely due to a delayed neuroendocrine feedback (Verasztó et al.,
2018). To determine whether NO mediates such a feedback, we repeated the
experiment in \emph{NOS}-mutant larvae. While we detected the initial
activation phase followed by inhibition, in homozygous
\emph{NOS}-mutants for both CRISPR alleles this was not followed by
delayed activation. Instead, calcium levels dropped to a low
steady-state level (Figure 4A, B). We thus identified a requirement for
NO signalling in the late-phase activation of cPRCs.
